Socioeconomic Determinants of Hate Crimes in the United States

### 2024-11-29 paper/paper.qmd
Modify paper
modify Data and Appendix section, add llm_usage and reference

---
title: "Socioeconomic Determinants of Hate Crimes in the United States"
subtitle: "A Bayesian Hierarchical Analysis"
author: 
  - Yuxuan Wei
thanks: "Code and data are available at: [https://github.com/wyx827/Hate_Crimes_Socioeconomic_Analysis]."
date: today
date-format: long
abstract: "This study provides the relationship between socioeconomic factors and hate crimes across U.S. states using a Bayesian hierarchical framework. We analyze a dataset capturing hate crime incidence alongside state-level socioeconomic indicators such as median income, unemployment rates, and income inequality. Results reveal that higher income inequality and unemployment rates are associated with elevated hate crime rates, while higher median income is inversely related. These findings emphasize the need for targeted socioeconomic interventions to mitigate hate crime prevalence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

# Library the required packages
library(dplyr)         
library(ggplot2)       
library(rstanarm)      
library(tidyr)        
library(arrow)         
library(readr)

@@

The remainder of this paper is structured as follows. @sec-data describes the dataset, variables, and preprocessing steps. Section 3 outlines the methodological framework, including the Bayesian hierarchical model. Section 4 presents the results, accompanied by tables and visualizations. Section 5 discusses the implications, limitations, and potential future directions of the analysis. Section 6 concludes with a summary of the key findings and recommendations.

@@

We use the statistical programming language R [@citeR], which provides a robust and flexible environment for statistical analysis and modeling. The data manipulation and cleaning tasks in this study are performed using [@tidyverse]. To visualize the variables and their relationships, we employ ggplot2 [@ggplot2], known for its elegant and versatile approach to graphical representation.

The Bayesian hierarchical model is implemented using the rstanarm package [@rstanarm], which facilitates applied Bayesian regression modeling in R. Additionally, the arrow package [@arrow] is used to handle efficient reading and writing of Parquet files, ensuring smooth data processing workflows.

Our dataset is sourced from FiveThirtyEight [@hatecrimesdata], which integrates hate crime statistics with state-level socioeconomic variables, providing a comprehensive basis for the analysis.

## Measurement

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

@@

Talk way more about it.

@@

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

@@

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

@@

@@

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

>>>>

---
title: "Socioeconomic Determinants of Hate Crimes in the United States"
subtitle: "A Bayesian Hierarchical Analysis"
author: 
  - Yuxuan Wei
thanks: "Code and data are available at: [https://github.com/wyx827/Hate_Crimes_Socioeconomic_Analysis]."
date: today
date-format: long
abstract: "This study provides the relationship between socioeconomic factors and hate crimes across U.S. states using a Bayesian hierarchical framework. We analyze a dataset capturing hate crime incidence alongside state-level socioeconomic indicators such as median income, unemployment rates, and income inequality. Results reveal that higher income inequality and unemployment rates are associated with elevated hate crime rates, while higher median income is inversely related. These findings emphasize the need for targeted socioeconomic interventions to mitigate hate crime prevalence."
format: pdf
math: true
number-sections: true
bibliography: references.bib
---

@@

# Library the required packages
library(dplyr)         
library(ggplot2)       
library(rstanarm)      
library(tidyr)        
library(arrow)         
library(readr)
library(corrplot)

@@

The remainder of this paper is structured as follows. @sec-data describes the dataset, variables, and preprocessing steps. @sec-modelsetup outlines the methodological framework, including the Bayesian hierarchical model. @sec-results  presents the results, accompanied by tables and visualizations. @sec-discussion discusses the implications, limitations, and potential future directions of the analysis. @sec-Appendix concludes with a summary of the key findings and recommendations.

@@

This study employs R [@citeR] for cleaning and analyzing the dataset, leveraging libraries such as `dplyr` [@tidyverse], `ggplot2` [@ggplot2], `corrplot` [@corrplot], and `arrow` [@arrow]. The dataset, sourced from FiveThirtyEight [@hatecrimesdata], investigates the relationships between socioeconomic variables and hate crime prevalence across the United States. It aggregates data for all 50 states and the District of Columbia.

After preprocessing, which involved removing missing values, standardizing variables, and ensuring data consistency, the final dataset includes **51** observations for **7** predictors: `median_income`, `unemployment_rate`, `metro_area_population_share`, `high_school_education_share`, `gini_index`, and `non_white_population_share`. The outcome variable, `hate_crimes_per_100k_splc`, represents hate crime prevalence reported by the SPLC. Together, these variables form the foundation for examining socioeconomic influences on hate crime rates.

---

## Measurement and Considerations

Understanding how socioeconomic conditions translate into measurable data is vital for interpreting the results. Below, we discuss the key measurement aspects and their implications.

- **Phenomena Representation**: Hate crime incidents, inherently complex societal phenomena, are represented in the dataset as `hate_crimes_per_100k_splc`, which quantifies the number of hate crimes reported per 100,000 residents in each state. This measure was derived from state-level aggregates reported by the SPLC, ensuring consistency across regions.

- **Data Collection Context**: Socioeconomic variables such as income, unemployment, and education are proxies for broader economic and social conditions. These data were collected from credible government sources and harmonized by FiveThirtyEight. The inclusion of standardized metrics, such as `gini_index` (inequality measure) and `metro_area_population_share`, ensures comparability across states despite regional disparities.

- **Challenges in Measurement**: Hate crime data, while critical, is subject to biases in underreporting and differences in enforcement. States with higher transparency or stricter enforcement may report higher rates, not necessarily indicating a greater prevalence. Similarly, socioeconomic indicators such as `non_white_population_share` could interact with other unmeasured factors, complicating causal interpretations.

- **Validation and Limitations**: Standardization ensures data comparability, but reliance on a single year of data limits trend analysis. This temporal snapshot captures static relationships, not dynamics over time. Additionally, state-level aggregation masks within-state heterogeneity, such as urban vs. rural differences in socioeconomic conditions and hate crime reporting.

---

## Variables of Interest

This section introduces the dataset's key variables, including detailed descriptions, visualizations, and insights into their relevance. Visualizations illustrate distributions, relationships, and variability to contextualize their potential influence on hate crime rates.

### Outcome Variable: Hate Crimes Per 100k SPLC

The outcome variable, representing the number of hate crimes per 100,000 residents as reported by the SPLC, its distribution, shown in @fig-hate-crimes, exhibits a skewed distribution. Most states report hate crime rates between 0.1 and 0.5 per 100,000 residents. However, the presence of an outlier with a much higher rate highlights significant regional disparities. This variation underscores the importance of investigating contributing factors, such as socioeconomic, demographic, and policy-related influences, to understand the drivers of hate crimes.

```{r}
#| label: fig-hate-crimes
#| fig-cap: Distribution of hate crimes per 100,000 residents, as reported by the SPLC, illustrating state-level variation. Most states report between 0.1 and 0.5 hate crimes per 100,000 residents, with a notable outlier indicating a disproportionately higher hate crime rate.
#| echo: false
#| warning: false
#| message: false

# Determine outlier values for better context
max_hate_crimes <- max(analysis_data$hate_crimes_per_100k_splc, na.rm = TRUE)

ggplot(analysis_data, aes(x = hate_crimes_per_100k_splc)) +
  geom_histogram(binwidth = 0.1, fill = "skyblue", color = "black", alpha = 0.7) +
  geom_vline(
    xintercept = mean(analysis_data$hate_crimes_per_100k_splc, na.rm = TRUE),
    color = "red", linetype = "dashed", label = "Mean"
  ) +
  geom_text(
    aes(
      x = mean(analysis_data$hate_crimes_per_100k_splc, na.rm = TRUE),
      y = max(table(cut(analysis_data$hate_crimes_per_100k_splc, breaks = 10))) - 1,
      label = paste0("Mean: ", round(mean(analysis_data$hate_crimes_per_100k_splc, na.rm = TRUE), 2))
    ),
    color = "red", vjust = 1
  ) +
  geom_text(
    aes(
      x = max_hate_crimes,
      y = 1,
      label = paste0("Outlier: ", round(max_hate_crimes, 2))
    ),
    color = "blue", vjust = -1
  ) +
  labs(
    title = "Distribution of Hate Crimes Per 100,000 Residents",
    x = "Hate Crimes Per 100,000 Residents",
    y = "Count of States"
  ) +
  theme_minimal()

@@

@@

### Median Income

Median income represents the annual income of the median household in each state, serving as a proxy for economic prosperity. As shown in @fig-median-income,  shows the distribution of median household income across states. There is a relatively normal distribution with a mean of $54,802 and a median of $54,310. The narrow spread suggests a clustering of household incomes around this value, with limited variation across states. This clustering might indicate consistency in income policies or similar economic opportunities in various regions of the United States.

```{r}
#| label: fig-median-income
#| fig-cap: Distribution of median household income across states, illustrating the variation in income levels. The mean household income is $54,802, while the median is slightly lower at $54,310, indicating a slightly right-skewed distribution. The standard deviation of $9,255 highlights variability across states.
#| echo: false
#| warning: false
#| message: false

ggplot(analysis_data, aes(x = median_income)) +
  geom_histogram(
    binwidth = 5000, 
    fill = "#3498db", 
    color = "#2980b9", 
    alpha = 0.7
  ) +
  geom_density(
    aes(y = after_stat(count)), 
    color = "#2c3e50", 
    fill = "#34495e", 
    alpha = 0.2
  ) +
  geom_vline(
    xintercept = c(mean(analysis_data$median_income), 
                   median(analysis_data$median_income)),
    linetype = c("dashed", "dotted"),
    color = c("red", "green")
  ) +
  annotate(
    "text", 
    x = mean(analysis_data$median_income) + 5000, 
    y = max(table(cut(analysis_data$median_income, breaks = 20))),
    label = paste(
      "Mean: $", round(mean(analysis_data$median_income), 0), 
      "\nMedian: $", round(median(analysis_data$median_income), 0),
      "\nSD: $", round(sd(analysis_data$median_income), 0)
    )
  ) +
  labs(
    title = "Distribution of Median Household Income",
    subtitle = paste("Number of States:", nrow(analysis_data)),
    x = "Median Income (USD)",
    y = "Count of States"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

### Unemployment Rate

The unemployment rate measures the percentage of individuals unemployed on a seasonal basis. As shown in @fig-unemployment, it shows an extremely narrow spread, with the mean and median unemployment rates both around 0.05%. This low variability suggests that unemployment rates are consistently low across states. The majority of states fall within a very small range of unemployment, which implies effective employment policies or consistent labor market dynamics nationwide.

```{r}
#| label: fig-unemployment
#| fig-cap: Distribution of unemployment rates among states. The unemployment rate is centered around 0.05%, showing very low variability among states, suggesting uniform employment opportunities across regions.
#| echo: false

# Compute the density data
density_data <- density(analysis_data$unemployment_rate, na.rm = TRUE)

# Find the leftmost and rightmost points where the density is non-zero
left_point <- min(density_data$x[density_data$y > 0])
right_point <- max(density_data$x[density_data$y > 0])

ggplot(analysis_data %>% filter(unemployment_rate >= 0 & unemployment_rate <= 10), 
       aes(x = unemployment_rate)) +
  # Histogram with appropriate binwidth
  geom_histogram(
    binwidth = 0.5, 
    fill = "#2ecc71", 
    color = "#27ae60", 
    alpha = 0.7,
  ) +
  # Density plot scaled to match histogram counts
  geom_density(
    aes(y = after_stat(density * nrow(analysis_data) * 0.5)), 
    color = "#2c3e50", 
    fill = "#2980b9", 
    alpha = 0.2
  ) +
  annotate(
    "text", 
    x = left_point - 0.15, 
    y = 50,
    label = paste("Left:", round(left_point, 3)), 
    color = "red"
  ) +
  annotate(
    "text", 
    x = right_point + 0.2, 
    y = 50, 
    label = paste("Right:", round(right_point, 3)), 
    color = "blue"
  ) +
  # Vertical lines for mean and median
  geom_vline(
    xintercept = c(mean(analysis_data$unemployment_rate, na.rm = TRUE), 
                   median(analysis_data$unemployment_rate, na.rm = TRUE)),
    linetype = c("dashed", "dotted"),
    color = c("red", "blue")
  ) +
  # Annotating statistical summary at a clearer location
  annotate(
    "text", 
    x = mean(analysis_data$unemployment_rate, na.rm = TRUE) + 1,  # Adjust x position
    y = max(table(cut(analysis_data$unemployment_rate, breaks = 20))) * 0.8,  # Adjust y position
    label = paste(
      "Mean: ", round(mean(analysis_data$unemployment_rate, na.rm = TRUE), 3), "%",
      "\nMedian: ", round(median(analysis_data$unemployment_rate, na.rm = TRUE), 3), "%",
      "\nSD: ", round(sd(analysis_data$unemployment_rate, na.rm = TRUE), 2), "%"
    ),
    hjust = 1,  # Adjust horizontal alignment
    vjust = -1,  # Adjust vertical alignment
    color = "black",
    size = 5
  ) +
  # Labels and title
  labs(
    title = "Distribution of Unemployment Rates",
    subtitle = paste("Number of States:", nrow(analysis_data)),
    x = "Unemployment Rate (%)",
    y = "Count of States"
  ) +
  # Minimal theme for clarity
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

### Metro Area Population Share

Metro area population share reflects the proportion of a state's population residing in metropolitan areas, which means states with higher urbanization levels may experience unique social dynamics influencing hate crime prevalence. As shown in @fig-metro-share, it shares reveals that the majority of states have a population share in metro areas concentrated around 0.77, with a standard deviation of 0.2. This suggests that many states have a high proportion of their populations in metropolitan areas, reflecting the urban-centric population distribution and possibly highlighting urbanization trends in the U.S.
```{r}
#| label: fig-metro-share
#| fig-cap: Distribution of metro area population share for states, showing the extent to which populations are concentrated in metropolitan areas. Most states have metro area shares clustered around 0.77, indicating a high proportion of urbanized population.
#| echo: false

ggplot(analysis_data, aes(x = metro_area_population_share)) +
  geom_histogram(
    binwidth = 5, 
    fill = "#f39c12", 
    color = "#d35400", 
    alpha = 0.7
  ) +
  geom_density(
    aes(y = after_stat(count)), 
    color = "#2c3e50", 
    fill = "#e67e22", 
    alpha = 0.2
  ) +
  # Annotate left point
  annotate(
    "text", 
    x = min(analysis_data$metro_area_population_share) - 0.25,   # Adjust horizontal position
    y = 50,                                                     # Adjust vertical position
    label = paste("Left:", round(min(analysis_data$metro_area_population_share), 3)), 
    color = "red",
    size = 4
  ) +
  # Annotate right point
  annotate(
    "text", 
    x = max(analysis_data$metro_area_population_share) + 0.45,   # Adjust horizontal position
    y = 50,                                                     # Adjust vertical position
    label = paste("Right:", round(max(analysis_data$metro_area_population_share), 3)), 
    color = "blue",
    size = 4
  ) +
  geom_vline(
    xintercept = c(mean(analysis_data$metro_area_population_share), 
                   median(analysis_data$metro_area_population_share)),
    linetype = c("dashed", "dotted"),
    color = c("red", "green")
  ) +
  annotate(
    "text", 
    x = mean(analysis_data$metro_area_population_share) + 5, 
    y = max(table(cut(analysis_data$metro_area_population_share, breaks = 20))),
    label = paste(
      "Mean: ", round(mean(analysis_data$metro_area_population_share), 2), "%",
      "\nMedian: ", round(median(analysis_data$metro_area_population_share), 2), "%",
      "\nSD: ", round(sd(analysis_data$metro_area_population_share), 1)
    ),
    hjust = 1,  # Adjust horizontal alignment
    vjust = -1  # Adjust vertical alignment
  ) +
  labs(
    title = "Metro Area Population Share Distribution",
    subtitle = paste("Number of States:", nrow(analysis_data)),
    x = "Metro Area Population Share (%)",
    y = "Count of States"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

### Gini Index

The Gini Index measures income inequality, with higher values indicating greater disparity. As shown in @fig-gini, it is centered around a mean of 0.456 and a median of 0.455. The relatively narrow spread in Gini values across states suggests that income inequality is consistent across the nation, with most states experiencing moderate levels of income disparity. This consistency may reflect similar economic structures or social policies.
```{r}
#| label: fig-gini
#| fig-cap: Distribution of Gini Index values across states, indicating income inequality. Most states exhibit a Gini Index value around 0.455, suggesting moderate levels of income inequality.
#| echo: false

ggplot(analysis_data, aes(x = gini_index)) +
  geom_histogram(
    binwidth = 0.02, 
    fill = "#9b59b6", 
    color = "#8e44ad", 
    alpha = 0.7
  ) +
  geom_density(
    aes(y = after_stat(count)), 
    color = "#2c3e50", 
    fill = "#673ab7", 
    alpha = 0.2
  ) +
  geom_vline(
    xintercept = c(mean(analysis_data$gini_index), 
                   median(analysis_data$gini_index)),
    linetype = c("dashed", "dotted"),
    color = c("red", "green")
  ) +
  annotate(
    "text", 
    x = mean(analysis_data$gini_index) + 0.05, 
    y = max(table(cut(analysis_data$gini_index, breaks = 20))),
    label = paste(
      "Mean: ", round(mean(analysis_data$gini_index), 3),
      "\nMedian: ", round(median(analysis_data$gini_index), 3),
      "\nSD: ", round(sd(analysis_data$gini_index), 2)
    ),
    hjust = 1,  # Adjust horizontal alignment
    vjust = -1  # Adjust vertical alignment
  ) +
  labs(
    title = "Distribution of Gini Index",
    subtitle = paste("Number of States:", nrow(analysis_data)),
    x = "Gini Index",
    y = "Count of States"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

### Non-White Population Share

This variable represents the proportion of non-white residents in each state. Higher non-white population shares may correlate with higher hate crime rates due to increased exposure to discriminatory behavior. As shown in @fig-nonwhite-share, this distribution of non-white population share indicates that most states have around 30% non-white residents. This suggests a considerable level of ethnic diversity across the nation. The mean of 0.32 and median of 0.3 show a moderate concentration of non-white population, reflecting varying levels of demographic diversity in different states.
```{r}
#| label: fig-nonwhite-share
#| fig-cap: Distribution of non-white population shares across states. The majority of states have a non-white population share around 0.3, reflecting regional diversity levels.
#| echo: false

ggplot(analysis_data, aes(x = non_white_population_share)) +
  geom_histogram(
    binwidth = 5, 
    fill = "#e74c3c", 
    color = "#c0392b", 
    alpha = 0.7
  ) +
  geom_density(
    aes(y = after_stat(count)), 
    color = "#2c3e50", 
    fill = "#ff6b6b", 
    alpha = 0.2
  ) +
  # Annotate left point
  annotate(
    "text", 
    x = min(analysis_data$non_white_population_share) - 0.6,   # Adjust horizontal position
    y = 50,                                                    # Adjust vertical position
    label = paste("Left:", round(min(analysis_data$non_white_population_share), 3)), 
    color = "red",
    size = 4
  ) +
  # Annotate right point
  annotate(
    "text", 
    x = max(analysis_data$non_white_population_share) + 0.6,   # Adjust horizontal position
    y = 50,                                                    # Adjust vertical position
    label = paste("Right:", round(max(analysis_data$non_white_population_share), 3)), 
    color = "blue",
    size = 4
  ) +
  geom_vline(
    xintercept = c(mean(analysis_data$non_white_population_share), 
                   median(analysis_data$non_white_population_share)),
    linetype = c("dashed", "dotted"),
    color = c("red", "green")
  ) +
  annotate(
    "text", 
    x = mean(analysis_data$non_white_population_share) + 5, 
    y = max(table(cut(analysis_data$non_white_population_share, breaks = 20))),
    label = paste(
      "Mean: ", round(mean(analysis_data$non_white_population_share), 2), "%",
      "\nMedian: ", round(median(analysis_data$non_white_population_share), 2), "%",
      "\nSD: ", round(sd(analysis_data$non_white_population_share), 1)
    ),
    hjust = 1,  # Adjust horizontal alignment
    vjust = -1  # Adjust vertical alignment
  ) +
  labs(
    title = "Non-White Population Share Distribution",
    subtitle = paste("Number of States:", nrow(analysis_data)),
    x = "Non-White Population Share (%)",
    y = "Count of States"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

## Summary Statistics and Variable Relationships
The relationships between predictors and hate crimes were explored using correlation matrices, summarized in @fig-corr. The correlation heatmap summarizes the relationships between various variables, providing insight into potential predictors of hate crimes per 100k SPLC. Notable correlations include the positive relationship between high school education share and metro area population share (r = 0.57), and between Gini Index and unemployment rate (r = 0.49). The relationship between hate crimes per 100k SPLC and non-white population share is weak, suggesting limited direct influence of ethnic diversity on hate crime prevalence at the state level.
```{r}
#| label: fig-corr
#| fig-cap: Correlation heatmap of predictor variables, showing relationships among median household income, unemployment rate, metro area population share, high school education share, Gini index, and hate crimes per 100k SPLC. (income = median_income, unemployment = unemployment_rate, population = metro_area_population_share, education = high_school_education_share, Gini = gini_index, non_white_population = non_white_population_share, hate_crimes = hate_crimes_per_100k_splc, avg_hatecrimes = avg_hatecrimes_per_100k_fbi)
#| echo: false

# Load necessary libraries
library(corrplot)

# Rename the variables for clarity
renamed_data <- analysis_data %>%
  rename(
    income = median_income,
    unemployment = unemployment_rate,
    population = metro_area_population_share,
    education = high_school_education_share,
    Gini = gini_index,
    non_white_population = non_white_population_share,
    hate_crimes = hate_crimes_per_100k_splc,
    avg_hatecrimes = avg_hatecrimes_per_100k_fbi
  )

# Generate the correlation heatmap
corrplot(
  cor(renamed_data %>% select(-state), use = "complete.obs"), 
  method = "color", 
  addCoef.col = "black",   # Add correlation coefficients
  tl.col = "black",        # Color of the text labels
  number.cex = 0.6,        # Reduce the size of the correlation coefficient text
  tl.cex = 0.6,            # Adjust text label size
  tl.srt = 45,             # Rotate text labels for clarity
  diag = FALSE             # Remove diagonal correlations
)

@@

## Model Selection
To analyze the relationship between socioeconomic factors and hate crimes, a **Bayesian hierarchical model** was chosen. This model accounts for state-level heterogeneity by including random effects for states, allowing for variability in hate crimes that might be influenced by unobserved state-specific factors. The Bayesian framework also allows incorporating prior knowledge and managing uncertainty effectively.

Alternative models were considered, including:

1. **Logistic Regression**: This model was used for scaling hate crimes within the range (0, 1). However, logistic regression is primarily suited for binary or bounded outcomes and may not capture the full variability of continuous predictors in this dataset.

2. **Simple Linear Regression**: While linear regression can estimate the relationship between hate crimes and individual predictors, it does not accommodate hierarchical structures or address multilevel variability, making it less robust for this dataset.

The **Bayesian hierarchical model** was selected as it provides the flexibility to model the complex relationships between hate crimes and predictors while considering the state-level effects.

Model diagnostics, implementation details, insights from the model summary, limitations, and alternative models considered are comprehensively documented in [Appendix -@sec-model], ensuring transparency and replicability of the analysis.

## Bayesian Hierarchical Model Overview

The model predicts hate crimes per 100,000 residents (`hate_crimes_per_100k_splc`) using the following predictors:

- **Median Household Income (`median_income`)**: Captures the economic status of residents within each state.
- **Unemployment Rate (`unemployment_rate`)**: Measures the proportion of the population without jobs.
- **Metro Area Population Share (`metro_area_population_share`)**: Reflects the proportion of residents living in metropolitan areas.
- **High School Education Share (`high_school_education_share`)**: Indicates the proportion of residents with at least a high school diploma.
- **Gini Index (`gini_index`)**: Measures income inequality within each state.
- **Non-White Population Share (`non_white_population_share`)**: Reflects the demographic diversity within each state.
- **State (`state`)**: Included as a random effect to capture state-specific variability.

The model takes the form:

\[
\text{Hate Crimes}_i = \beta_0 + \beta_1 \cdot \text{Income}_i + \beta_2 \cdot \text{Unemployment}_i + \beta_3 \cdot \text{Population}_i + \beta_4 \cdot \text{Education}_i + \beta_5 \cdot \text{Gini}_i + \beta_6 \cdot \text{NonWhite}_i + u_i + \epsilon_i
\]

Where:

- \(\beta_0\) is the intercept term.
- \(\beta_1, \beta_2, \dots, \beta_6\) are coefficients for the fixed effects (predictors).
- \(u_i\) represents the random effects for state-level variations.
- \(\epsilon_i \sim \text{Normal}(0, \sigma^2)\) is the residual error term.

### Assumptions and Priors

Key assumptions for the Bayesian hierarchical model include:

1. **Linearity**: The relationship between predictors and the outcome is linear.
2. **Homoscedasticity**: The residuals have constant variance.
3. **Random Effects**: State-level random effects capture unobserved variability.
4. **Normality**: Errors are normally distributed.
5. **Priors**: Normal priors with mean 0 and standard deviation 2.5 are used for coefficients and intercepts, allowing for flexibility while avoiding extreme values.

The prior distributions are as follows:
\begin{align}
\beta_j &\sim \text{Normal}(0, 2.5), \quad \forall j
\end{align}

## Interpretation of Coefficients
The Bayesian hierarchical model estimates the relationship between socioeconomic predictors and hate crimes per 100,000 residents. Below is the interpretation of each coefficient:

### Fixed Effects

- **Intercept (\(\beta_0\))**: 
  The intercept represents the baseline predicted hate crimes per 100,000 residents when all predictors (median income, unemployment rate, metro area population share, high school education share, Gini index, and non-white population share) are at their mean or baseline values. This serves as the reference point for interpreting the effects of the predictors.

- **Median Household Income (\(\beta_1\))**: 
  This coefficient indicates how a $1,000 increase in median household income affects the predicted hate crimes per 100,000 residents. A negative coefficient suggests that higher income is associated with fewer hate crimes, potentially reflecting improved economic stability and reduced socioeconomic tension.

- **Unemployment Rate (\(\beta_2\))**: 
  This coefficient captures the effect of a 1% increase in the unemployment rate on the predicted hate crimes per 100,000 residents. A positive coefficient implies that higher unemployment is associated with more hate crimes, possibly due to heightened social stress and economic insecurity.

- **Metro Area Population Share (\(\beta_3\))**: 
  This coefficient represents the effect of a 1% increase in the proportion of residents living in metropolitan areas on the predicted hate crimes. A positive coefficient could indicate that urban areas, with their higher population density, may experience more hate crimes due to greater diversity and interpersonal interactions.

- **High School Education Share (\(\beta_4\))**: 
  This coefficient reflects how a 1% increase in the proportion of residents with at least a high school diploma impacts the predicted hate crimes. A negative coefficient suggests that higher education levels are associated with fewer hate crimes, highlighting the role of education in promoting tolerance and reducing prejudice.

- **Gini Index (\(\beta_5\))**: 
  The Gini index measures income inequality. This coefficient shows how a 0.01 increase in the Gini index affects hate crimes. A positive coefficient indicates that higher income inequality correlates with more hate crimes, suggesting that economic disparities may contribute to social tensions.

- **Non-White Population Share (\(\beta_6\))**: 
  This coefficient captures the effect of a 1% increase in the non-white population share on hate crimes. A positive coefficient may reflect increased targeting of minority groups in states with greater diversity.

### Random Effects

- **State-Level Effects (\(u_i\))**: 
  The random effects account for state-specific variability in hate crime rates that cannot be explained by the predictors. These effects capture unobserved factors unique to each state, such as cultural, political, or enforcement differences.

### Residual Variance (\(\sigma^2\))

The residual variance represents the variability in hate crimes not accounted for by the model. A smaller residual variance indicates that the model explains more of the variability in hate crimes.

---

## Example Interpretation

If the coefficient for **unemployment rate** (\(\beta_2\)) is \(0.25\), then a 1% increase in unemployment rate would lead to an increase of \(0.25\) hate crimes per 100,000 residents, holding all other predictors constant. Similarly, if the coefficient for **median household income** (\(\beta_1\)) is \(-0.05\), then a $1,000 increase in income would reduce hate crimes by \(0.05\) per 100,000 residents.

## Implications

The coefficients provide insights into how socioeconomic factors impact hate crimes. For example:

- Policies targeting economic inequality (reducing the Gini index) and increasing educational attainment may help reduce hate crimes.
- Urbanization and population diversity may require targeted interventions to address potential tensions.

These interpretations guide the understanding of how predictors influence hate crimes and inform potential policy recommendations.

## Model Justification
The following are interpretations of the predictors' coefficients from the Bayesian hierarchical model:

- **Median Income**: The coefficient indicates how changes in state median income (in $1,000 units) affect hate crimes per 100,000 residents. A negative coefficient suggests that higher income reduces hate crimes, reflecting the protective effect of economic stability.

- **Unemployment Rate**: This coefficient shows how a 1% increase in unemployment affects hate crimes. A positive and statistically significant coefficient suggests that higher unemployment correlates with more hate crimes, likely due to increased social stress.

- **Metro Area Population Share**: This coefficient reflects the effect of a 1% increase in metro area population share on hate crimes. A positive coefficient may indicate that more densely populated or urbanized areas experience slightly higher hate crimes, potentially due to greater diversity and social interactions.

- **High School Education Share**: The coefficient represents the impact of a 1% increase in high school education attainment. A negative coefficient suggests that higher education levels reduce hate crimes, possibly by promoting tolerance and reducing prejudice.

- **Gini Index**: The coefficient measures the effect of a 0.01 increase in the Gini index (income inequality). A positive coefficient suggests that higher income inequality is associated with more hate crimes, emphasizing the role of economic disparities in fostering tension.

- **Non-White Population Share**: This coefficient captures the effect of a 1% increase in non-white population share. A positive coefficient may indicate that states with larger non-white populations experience more reported hate crimes, potentially reflecting targeted discrimination or diversity-related tensions.

```{r}
#| echo: false
#| label: tbl-summary-model
#| tbl-cap: Coefficient estimates from the Bayesian hierarchical model. The table shows the relationship between predictors and hate crimes per 100,000 residents (SPLC).
#| warning: false
#| message: false

library(rstanarm)
library(here)

# Load the Bayesian model
bayesian_model <- readRDS(here::here("models/final_bayesian_model.rds"))

# Extract the posterior draws as a matrix
posterior_draws <- as.matrix(bayesian_model)

# Calculate the posterior summaries
posterior_summary <- apply(posterior_draws, 2, function(x) {
  c(
    Mean = mean(x),
    `Std. Error` = sd(x),
    `2.5% CI` = quantile(x, 0.025),
    `97.5% CI` = quantile(x, 0.975)
  )
})

# Convert to a data frame
posterior_summary_df <- as.data.frame(t(posterior_summary))

# Rename columns
colnames(posterior_summary_df) <- c("Mean", "Std. Error", "2.5% CI", "97.5% CI")

# Add Parameter names as a column (if missing)
posterior_summary_df$Parameter <- rownames(posterior_summary_df)

# Filter and rearrange columns
posterior_summary_df <- posterior_summary_df[, c("Parameter", "Mean", "Std. Error", "2.5% CI", "97.5% CI")]

# Display as a formatted table
knitr::kable(posterior_summary_df, digits = 3, caption = "Posterior Estimates for Model Coefficients")
```

@@

##  Outcome Variable: Hate Crimes Per 100k SPLC
The histogram @fig-hate-crimes visualizes the distribution of hate crimes per 100,000 residents across U.S. states, as reported by the SPLC.

- **Concentration of Values**: 
1. The majority of states report hate crime rates below 0.5 per 100,000 residents.
2. A substantial portion of states cluster between 0.0 and 0.3 hate crimes per 100,000 residents, indicating that low rates are common nationwide.

- **Skewed Distribution**: The distribution is right-skewed, with a long tail extending toward higher hate crime rates. This suggests that while most states experience relatively low hate crime rates, a few states report significantly higher rates.

- **Outliers**: There is at least one state with a hate crime rate exceeding 1.0 per 100,000 residents, marking it as an outlier with unusually high prevalence. (State:District of Columbia, with hate crime rate **1.522302**)

```{r}
#| echo: false
#| eval: false
#| warning: false
#| message: false

# Identify the outlier state
outlier_state <- analysis_data %>%
  filter(hate_crimes_per_100k_splc > 1.0) %>%
  select(state, hate_crimes_per_100k_splc)

# Print the outlier state(s)
print(outlier_state)
```

- **Implications for Analysis**:
1. The variability in hate crime rates emphasizes the importance of analyzing socioeconomic factors to understand why certain states report higher rates.
2. The skewness and presence of outliers highlight the need for robust statistical techniques to ensure results are not overly influenced by extreme values.

This visualization provides a foundational understanding of the outcome variable, setting the stage for exploring relationships with predictor variables like income inequality, unemployment, and urbanization.

@@

## Diagnostics and Validation

### Trace Plot for MCMC Chains
